---
title: "R4MS Workshop 2 - Final Exercise"
author: "Ben Cresswell"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

# Introduction

Here we are going to wrangling in messy data on the Ross River estuary fish surveys and produce final clean tables and figure.
We have 4 separate files to wrangle in and use

# Housekeeping

```{r housekeeping}
#| message: true
#| warning: false
rm(list = ls())
```

# Packages

```{r}
library(tidyverse) # Inc dplyr, ggplot2, lubridate etc
library(readxl) # For reading in excel files
```

# Load files

# Main dataset
The catch data is split across 4 tabs in one excel file
Load indivudally
```{r}
library(readxl)
library(tidyverse)
catch1 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = 1)
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)

catch_data <- bind_rows(catch1, catch2, catch3, catch4)
```

# Alternative method for loading files
```{r}
file_path <- "../data/workshop2/estuary_catch_log.xlsx"
catch_data2 <- excel_sheets(file_path) |>
  map_dfr(~read_excel(path = file_path, sheet = .x))
```

# Examine catch data
```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```
# need to fix site and place in reasonable order, but we can load metadata to ensure consistency
# need to fix species names, but we can load taxonomic dictionary to translate to scientific names

# Load metadata
```{r}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")

# Inspect
head(metadata)
glimpse(metadata)
```

# Site names are all over the place and need standardised for metadata
```{r}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |>
  # convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```

# Recheck
```{r}
catch_data |> 
  str()
# catch_data |> 
#   distinct(site)
glimpse(metadata)
```


# Load Taxonomic Dictionary
```{r}
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")
head(species_dict)
```

# Load sensor data
```{r}
sensor_raw <- read_csv("../data/workshop2/estuary_sonde_data.csv")
```

# Check sensor data
```{r}
head(sensor_raw)
str(sensor_raw)
glimpse(sensor_raw)
summary(sensor_raw)
```

### Clean up sensor data
```{r}
sensor_data <- 
  sensor_raw |>
  mutate(
    # Break down the messy character string into a proper datetime object
    datetime = dmy_hm(timestamp),
    # Convert the -999.0 hardware error to a true NA value
    clean_turbidity = na_if(turbidity, -999),
    # Convert negative salinit values to NA
    clean_salinity = ifelse(salinity < 0, NA, salinity)
  )
```

# Standardise variable in the catch log to match dictionaries
```{r}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_"),
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")
  )
```

# Check common names in catch log match those in the taxonomic dictionary
```{r}
catch_data |> 
  distinct(species)
species_dict |>
  distinct(common_name)
```

# Taxonomic translation (common to scientific names)
```{r}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(species_dict, by = join_by(common_name)) |> 
  select(site, date, common_name, scientific_name, count) |>
  select(-common_name)
```


