---
title: "R4MS Workshop 2 - Final Exercise"
author: "Ben Cresswell"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = TRUE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

# Introduction


Here we are going to wrangle in messy data examining the relationship between water quality and predatory fish assemblages along the Ross River Estuary gradient to produce final clean tables and figure.
We have 4 separate files to use and wrangle in

# Housekeeping

```{r housekeeping}
#| message: true
#| warning: false
rm(list = ls())
```

# Packages

```{r}
library(tidyverse) # Inc dplyr, ggplot2, lubridate etc
library(readxl) # For reading in excel files
```

# Load files

# Main dataset
The catch data is split across 4 tabs in one excel file
Load indivudally
```{r}
catch1 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = 1)
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)

# Bind rows together
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
```

# Alternative method for loading files
```{r}
file_path <- "../data/workshop2/estuary_catch_log.xlsx"
catch_data2 <- excel_sheets(file_path) |>
  map_dfr(~read_excel(path = file_path, sheet = .x))
```

# Examine catch data
```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)

#need to check how many sites and species we have
catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)
```
# need to fix site and place in reasonable order, but we can load metadata to ensure consistency
# need to fix species names, but we can load taxonomic dictionary to translate to scientific names

# Load metadata
```{r}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")

# Inspect
head(metadata)
glimpse(metadata)
```

# Site names are all over the place and need standardised for both tables of data and metadata
```{r}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |>
  # convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```

# Recheck
```{r}
catch_data |> 
  str()
# catch_data |> 
#   distinct(site)
glimpse(metadata)
```

# Load Taxonomic Dictionary
```{r}
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")
head(species_dict)
```

# Check common names in catch log match those in the taxonomic dictionary
```{r}
catch_data |> 
  distinct(species)
species_dict |>
  distinct(common_name)
```

# Standardise species names in catch log to match taxonomic dictionary
```{r}
catch_data <- catch_data |>
  mutate(
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")
  )
```

# Taxonomic translation (common to scientific names) by creating new variable and joining tables
```{r}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(species_dict, by = join_by(common_name)) |> 
  select(site, date, common_name, scientific_name, count) |>
  select(-common_name)

# Check to see if data is fixed
glimpse(catch_data)
```

# Load sensor data
```{r}
# Code can be written in separate scripts if it is more complex and you don't want the md file to be extra long, it then can be sourced in so the code runs fully in the md file without all the extra work shown - can still upload separate script to GitHub so the script is accessible if anyone wants to see how the code works.
source(here::here("code/diw_sensor.R"))
```

```{r join all data}
library(tidyverse)
catch_data_final <- catch_data |>
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sensor_daily, by = join_by(site, date))
```


# Standardise variable in the catch log to match dictionaries
```{r}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_"),
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")
  )
```


```{r}

```
# Wrangle final dataset

```{r wrangle-final-zeros}
catch_data_final <- 
catch_data_final |> 
  complete(
    nesting(site, date, 
           # lat, lon, 
            zone, mean_temp, mean_salinity, mean_turbidity),
    scientific_name
  ) |>
  mutate(
    # Convert the resulting NA counts to 0
    count = coalesce(count, 0)
  )
```


## Now start thinking about hypotheses and what might be going on in the ecology of these species in this ecosystem

### first check that there is an actual difference in env vars among zones

#### Salinity
```{r}
# table
sal_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
   # sem_salinity = sd(mean_salinity) / sqrt(n()),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )

# plot
sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
```


Definitely differences in salinity, And we know there has been work on some of these species about responses to salinity gradients.

Let's check the other two env vars too

### Turbidity
```{r}
turb_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_turbidity = sd(mean_turbidity, na.rm = TRUE),
    median_turbidity = median(mean_turbidity, na.rm = TRUE),
    .groups = "drop"
  )

turb_sum |>
  ggplot(aes(x = zone, y = median_turbidity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_turbidity - sd_turbidity, ymax = median_turbidity + sd_turbidity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Turbidity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Turbidity (NTU)")
```


### And temperature
```{r}
temp_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_temperature = sd(mean_temp, na.rm = TRUE),
    median_temperature = median(mean_temp, na.rm = TRUE),
    .groups = "drop"
  )

temp_sum |>
  ggplot(aes(x = zone, y = median_temperature)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_temperature - sd_temperature, ymax = median_temperature + sd_temperature), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Temperature by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Temperature (°C)")

```


OK so we'll continue with salinity as our main env var of interest, but we know that there are also differences in turbidity and temperature across the zones, which could also be influencing fish abundance patterns. We can discuss this in the interpretation of our results and consider potential confounding effects of these other environmental variables in our discussion.


#### Remake salinity plot nicely for publication
```{r}
#| fig-cap: "Salinity Gradient across zones. Ross River Estuary System, Townsville, NQ, Australia May 2026."

sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
  
```

# Now we can turn to patterns in fish abundance


# Calculate summary stats 
```{r}
fish_stats <- catch_data_final |>
  group_by(zone, scientific_name) |>
  summarise(
    mean_catch = mean(count),
    sem_catch = sd(count) / sqrt(n()),
    .groups = "drop"
  )
```

# And plot

```{r}
fish_stats |> 
  ggplot(aes(x = zone, y = mean_catch, fill = zone)) +
  geom_line(group = 1) +
  # add in points from master data
  geom_point(data = catch_data_final, aes(x = zone, y = count), 
             position = position_jitter(width = 0.2), alpha = 0.2, size = 1, inherit.aes = FALSE) +
  geom_point(position = "dodge", size = 4, pch = 23) +
  
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch), 
                position = position_dodge(0.9), width = 0.2) +
  
  facet_wrap(~ scientific_name, scales = "fixed") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey20"),
   # legend.position = "bottom"
    legend.position = "none"
  ) +
  labs(
    #title = "Predatory Fish Abundance across Estuarine Salinity Gradients",
    #subtitle = "Ross River Estuary System, May 2026",
    x = "Spatial Zone",
    y = "Abundance (per day)",
    fill = "Spatial Zone"
  )
```




