---
title: "Workshop2"
Author: "Natalie Wasinski"
output: html_document
---

# Advanced data wrangling: Extracting ecological signals from noisy systems

library(tidyverse)

# Tidying data

## Step 1: understand what each variable and observation actually means

## Step 2: resolve one of the two common problems with untidy data - one variable is spread across multiple columns or one observation is scattered across multiple rows (pivot data to fix these)

### tidyr: pivot_longer() to lengthen data and pivot_wider() to widen data

### Rules of tidy data: each variable must have its own column, each observation must have its own row, and each value must have its own cell

```{r 2.4 Lengthening Datasets}
billboard
#> # A tibble: 317 × 79
#>   artist       track               date.entered   wk1   wk2   wk3   wk4   wk5
#>   <chr>        <chr>               <date>       <dbl> <dbl> <dbl> <dbl> <dbl>
#> 1 2 Pac        Baby Don't Cry (Ke... 2000-02-26      87    82    72    77    87
#> 2 2Ge+her      The Hardest Part O... 2000-09-02      91    87    92    NA    NA
#> 3 3 Doors Down Kryptonite          2000-04-08      81    70    68    67    66
#> 4 3 Doors Down Loser               2000-10-21      76    76    72    69    67
#> 5 504 Boyz     Wobble Wobble       2000-04-15      57    34    25    17    17
#> 6 98^0         Give Me Just One N... 2000-08-19      51    39    34    26    26
#> # ℹ 311 more rows
#> # ℹ 71 more variables: wk6 <dbl>, wk7 <dbl>, wk8 <dbl>, wk9 <dbl>, ...

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", values_to = "rank"
  )
#> # A tibble: 24,092 × 5
#>    artist track                   date.entered week   rank
#>    <chr>  <chr>                   <date>       <chr> <dbl>
#>  1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#>  2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#>  3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#>  4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#>  5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#>  6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#>  #> # ℹ 24,082 more rows

# When data was lengthened, the weeks where a song wasn't on the charts became 'NA' as forced to exist because of the structure of the dataset. Ask pivot_longer to remove by adding values_drop_na = TRUE

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
 )
#> # A tibble: 5,307 × 5
#>   artist track                   date.entered week   rank
#>   <chr>  <chr>                   <date>       <chr> <dbl>
#> 1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#> 2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#> 3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#> 4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#> 5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#> 6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#> # ℹ 5,301 more rows

# far fewer rows than previous table, indicating that heaps of NAs were dropped.
# to convert some of our values from strings to numbers we can use mutate() and parse_number() to extract the number from the week column and convert it to a numeric value

# Create dataset
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)

#want to tidy data to have 3 variables: id, measurement (column), and value (cell values)
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
#> # A tibble: 6 × 3
#>   id    measurement value
#>   <chr> <chr>       <dbl>
#> 1 A     bp1           100
#> 2 A     bp2           120
#> 3 B     bp1           140
#> 4 B     bp2           115
#> 5 C     bp1           120
#> 6 C     bp2           125
```

```{r 2.5 Widening Datasets}
cms_patient_experience
#> # A tibble: 500 × 5
#>   org_pac_id org_nm                     measure_cd   measure_title   prf_rate
#>   <chr>      <chr>                      <chr>        <chr>              <dbl>
#> 1 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_1  CAHPS for MIPS...       63
#> 2 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_2  CAHPS for MIPS...       87
#> 3 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_3  CAHPS for MIPS...       86
#> 4 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_5  CAHPS for MIPS...       57
#> 5 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_8  CAHPS for MIPS...       85
#> 6 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_12 CAHPS for MIPS...       24
#> # ℹ 494 more rows

# To see the complete set of values for measure_cd and measure_title using distinct():
cms_patient_experience |> 
  distinct(measure_cd, measure_title)
#> # A tibble: 6 × 2
#>   measure_cd   measure_title                                                 
#>   <chr>        <chr>                                                         
#> 1 CAHPS_GRP_1  CAHPS for MIPS SSM: Getting Timely Care, Appointments, and In...
#> 2 CAHPS_GRP_2  CAHPS for MIPS SSM: How Well Providers Communicate            
#> 3 CAHPS_GRP_3  CAHPS for MIPS SSM: Patient's Rating of Provider              
#> 4 CAHPS_GRP_5  CAHPS for MIPS SSM: Health Promotion and Education            
#> 5 CAHPS_GRP_8  CAHPS for MIPS SSM: Courteous and Helpful Office Staff        
#> 6 CAHPS_GRP_12 CAHPS for MIPS SSM: Stewardship of Patient Resources

# pivot_wider() has the opposite interface to pivot_longer(): instead of choosing new colmn names we need to provide the existing columns that define the values and column name
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,values_from = prf_rate
  )
#> # A tibble: 500 × 9
#>   org_pac_id org_nm                   measure_title   CAHPS_GRP_1 CAHPS_GRP_2
#>   <chr>      <chr>                    <chr>                 <dbl>       <dbl>
#> 1 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          63          NA
#> 2 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          87
#> 3 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          NA
#> 4 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          NA
#> 5 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          NA
#> 6 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          NA
#> # ℹ 494 more rows
#> # ℹ 4 more variables: CAHPS_GRP_3 <dbl>, CAHPS_GRP_5 <dbl>, ...

# Still need to tell which columns have vlaues that uniquely idnetify each row to make data tidy
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
#> # A tibble: 95 × 8
#>   org_pac_id org_nm           CAHPS_GRP_1 CAHPS_GRP_2 CAHPS_GRP_3 CAHPS_GRP_5
#>   <chr>      <chr>                  <dbl>       <dbl>       <dbl>       <dbl>
#> 1 0446157747 USC CARE MEDICA...          63          87          86          57
#> 2 0446162697 ASSOCIATION OF ...          59          85          83          63
#> 3 0547164295 BEAVER MEDICAL ...          49          NA          75          44
#> 4 0749333730 CAPE PHYSICIANS...          67          84          85          65
#> 5 0840104360 ALLIANCE PHYSIC...          66          87          87          64
#> 6 0840109864 REX HOSPITAL INC          73          87          84          67
#> # ℹ 89 more rows
#> # ℹ 2 more variables: CAHPS_GRP_8 <dbl>, CAHPS_GRP_12 <dbl>


## Pivoting wider example ##
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

# Take the names fromt he measurement column and the values from the value column
df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
#> # A tibble: 2 × 4
#>   id      bp1   bp2   bp3
#>   <chr> <dbl> <dbl> <dbl>
#> 1 A       100   120   105
#> 2 B       140   115    NA

# New column names will be the unique values of measurement
df |> 
  distinct(measurement) |> 
  pull()
#> [1] "bp1" "bp2" "bp3"

#rows in the output are determined by all variables that aren't going into new names or values
df |> 
  select(-measurement, -value) |> 
  distinct()
#> # A tibble: 2 × 1
#>   id   
#>   <chr>
#> 1 A    
#> 2 B

# pivot_wider combines these results to generate empty dataframe
df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)
#> # A tibble: 2 × 4
#>   id    x     y     z    
#>   <chr> <lgl> <lgl> <lgl>
#> 1 A     NA    NA    NA   
#> 2 B     NA    NA    NA
```

```{r 2.6 Pivoting Exervises using Palmer Penguins}
# Plot variables together stacked into two columns, one for measurement name and one holding numerical values
library(palmerpenguins)

# Create a long version of the penguins dataset

penguins_long <- penguins |>
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement_type",
    values_to = "value"
  )
# View the result
head(penguins_long)


## visualize distributions of what ggplot2 expects for multi-panel faceting ##
penguins_long |>
  drop_na(value) |>
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )
# Produces 4 graphs with different x axes but the same y axis, and each graph is a different measurement type. Each graph has the distribution of values for that measurement type across the three species of penguins, with different colors for each species.


## Widening data ## ------------

# Calculate mean body mass for each species on each island
mass_summary <- penguins |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g))

head(mass_summary)
# still not tidy enough for publication presentation, widen to do this

mass_matrix <- mass_summary |>
  pivot_wider(
    names_from = island,
    values_from = mean_mass
  )

head(mass_matrix)
# names_from dictates which column procides new headers (islands) and values_from dictates which column populates the cells (mean mass). NA values automatically populate where specific species don't occur on island.
```

```{r 2.7 Separating and uniting data tables}
# Table 3 needs to split the rate column up into two variables, take the name of the column we want to split and the names of the columns we want to split into
table3 %>% 
  separate(rate, into = c("cases", "population"))
#> # A tibble: 6 × 4
#>   country      year cases  population
#>   <chr>       <int> <chr>  <chr>     
#> 1 Afghanistan  1999 745    19987071  
#> 2 Afghanistan  2000 2666   20595360  
#> 3 Brazil       1999 37737  172006362 
#> 4 Brazil       2000 80488  174504898 
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583

# if you want to use a specific character to separate a column, you can pass the character to the sep argument of separate()
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")

# Cases and population are listed as character types, convert them into numbers
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583


# Pass a vector of integers to sep - seperate() will interpret integers as positions to split at. When using integers to separate strings, the length of sep should be one less than the number of names in into. Can use this arrangement to separate the last two digits of each year (makes this specific data less tidy but can be useful in other contexts)
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
#> # A tibble: 6 × 4
#>   country     century year  rate             
#>   <chr>       <chr>   <chr> <chr>            
#> 1 Afghanistan 19      99    745/19987071     
#> 2 Afghanistan 20      00    2666/20595360    
#> 3 Brazil      19      99    37737/172006362  
#> 4 Brazil      20      00    80488/174504898  
#> 5 China       19      99    212258/1272915272
#> 6 China       20      00    213766/1280428583


## Unite(): inverse of separate() - combines multiple columns into a single column ##
table5 %>% 
  unite(new, century, year, sep = "")
#> # A tibble: 6 × 3
#>   country     new   rate             
#>   <chr>       <chr> <chr>            
#> 1 Afghanistan 1999  745/19987071     
#> 2 Afghanistan 2000  2666/20595360    
#> 3 Brazil      1999  37737/172006362  
#> 4 Brazil      2000  80488/174504898  
#> 5 China       1999  212258/1272915272
#> 6 China       2000  213766/1280428583

# Here we need to add sep ="" because we don’t want any separator (the default is to add an underscore _)

```

```{r 2.8 Wrangling strings and dates}

## Standardizing text with stringer ## -----
# all functions in the stringr package begin with str_ making them easy to find using autocomplete

library(tidyverse)

# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)

# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
    # 2. Replace any spaces with underscores
    #site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_"),
    # 3. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_trim(site_clean)
  )

print(clean_sites)
# Still provides 2 variations of same site name because order matters, need to switch lines 2 and 3

## CORRECT VERSION ## ---
library(tidyverse)

# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)

# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_trim(site_clean),
    # 3. Replace any spaces with underscores
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_")
  )

print(clean_sites)


## Mastering time with lubridate ## ------

library(lubridate)

# Parsing different date formats
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

# R now recognizes these as identical Date objects
date_1 == date_2

# For exact timestamps:
# A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)

```

```{r 2.9 Relational data: Joining tables}
# Set up two tables to join
# Table 1: Biological observation data
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"),
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)

# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP", "WP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "Other Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)

## The workhorse: left_ and right_join() ## ----

# left_join()keeps all the rows from your left table (observations) and matches data from your right table (site_metadata) wherever the key matches. If a match is missing, it fills the gaps with NA.

# Joining metadata to our observations
joined_data <- observations |>
  left_join(site_metadata, by = join_by(site_code))

print(joined_data)

## The strict filter: inner_join() ## ----

# An inner_join() only keeps rows that have complete matches in both tables. Use this when you only want to analyze data that has complete spatial or environmental context.

matched_data <- observations |>
  inner_join(site_metadata, by = join_by(site_code))
glimpse(matched_data)

## The detective: anti_join() ## ----

# Filters left table to only show rows that do not have a match in the right table. Helps to find out which site codes are misspelled or missing from your metadata table.

# Which observations are missing from our metadata dictionary?
missing_context <- observations |>
  anti_join(site_metadata, by = join_by(site_code)) 

print(missing_context)
```

```{r 2.10 Handling missing values}

## Converting legacy errors with na_if() ## ----

# Raw data from an old temperature logger
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)) # -999 is a known sensor error code

# Convert the -999 error codes to true NA values
fixed_logger <- logger_data |>
  mutate(temp_c = na_if(temp_c, -999))

print(fixed_logger)
# Important error to clean at the beginning of analysis

## Replacing missing values with coalesce() ## ----

# When NA is actually supposed to be a zero number value

# Simulate some count data
shark_counts <- tibble(
  site = c("Reef_A", "Reef_B", "Reef_C"),
  shark_count = c(3, NA, 5)) # The NA here actually means 0 sharks were seen

# Replace NA with 0 
shark_fixed <- shark_counts |>
  mutate(shark_count = coalesce(shark_count, 0))

print(shark_fixed)

## Special case of NaN (not a number) ## ----

#NaN means this value is mathematically impossible, rather than we don't know what the value is - most often when performing a math operation with indeterminate result (like 0/0 or Inf - Inf)

cpue_data <- tibble(
  site = c("Bay_1", "Bay_2"),
  catch = c(10, 0),
  effort_hours = c(2, 0))

# Calculating CPUE (catch / effort)
cpue_calc <- cpue_data |>
  mutate(
    cpue = catch / effort_hours)

print(cpue_calc)

## Implicit missing values and the zero-data framework: complete() ## ----

# only lists what was caught
raw_catch <- tibble(
  site = c("Reef_1", "Reef_1", "Reef_2"),
  species = c("Pmaculatus", "Pleopardus", "Pmaculatus"),
  count = c(5, 2, 8))
# implied reef_2 was surveyed but nothing found there

# use tidyr::complete() to build a zero-data framework. Finds all unique combinations of the variables you supply, explicitly creates rows for the missing combinations, and allows you to fill the resulting NAs with 0

# Force inclusion of hidden zero-catch data
full_catch_matrix <- raw_catch |>
  complete(site, species, fill = list(count = 0))

print(full_catch_matrix)

## The nuclear option: drop_na() ## ----

# If an NA represents a failed instrument reading, and that reading is the core response variable of your analysis, you may have no choice but to remove the entire row
sensor_log <- tibble(
  day = 1:4,
  salinity = c(35.2, 35.1, NA, 35.3)
)

# Remove any row where salinity is missing
clean_log <- sensor_log |>
  drop_na(salinity)

print(clean_log)
```

```{r 2.11 Practical Exercises}

```

```{r 2.12 Keystone Exercise}
# Try on own but will Work on at beginning of next workshop #
```

